NumPy---Array Programming in Python

### 2020-01-25 body.tex
Revise discussion

Exposing array programming primitives, as well as the surrounding ecosystem of
tools, in Python---an interpreted language---creates an ideal environment for
interactive, exploratory data analysis where users may iteratively inspect,
manipulate, and visualize their data \cite{perez2007ipython}.

@@

% aka why is this so successful
% some sense of ongoing work / future directions

@@

NumPy has subsequently seen prolific adoption, with millions of users,
first from research and later also industry.  It spurred the
development of the larger scientific Python ecosystem and heralded the
current era of wide-spread use of Python for scientific computing. The
reasons for its success are certainly complex, but we argue that they
are broadly of \emph{practical}, \emph{philosophical},
\emph{social}, and \emph{technical} origins.

% Fortran / C
% A few sentences on the history of array programming, referencing APL and Fortran, are missing.

The \emph{practical} appeal of NumPy and its ecosystem is apparent
when you consider the following description of how APL users
work---one which
easily applies to scientific Python users too \cite{kromberg2007apl}:
\begin{quote}
Users of APL tend to ``live in their data''. Interactive interpreters allow them
to inspect, almost feel their  way forward, discovering successful snippets
of code through experimentation and collecting them
into imperative or functional programs according to taste.
\end{quote}

This modality of interactive scientific exploration would develop
further and eventually culminate in the Jupyter ecosystem of tools \fixme{citation needed}.

NumPy also benefited from being written in Python, a
general-purpose language well-suited to standard programming tasks such as
cleaning data, interacting with web resources, and parsing text.
Adding fast array operations and linear algebra allows scientist to do all
their work with within a single language---and one that has the advantage of
being famously easy to learn and teach.  In fact, Python's adoption as
a primary learning language in many universities \fixme{citation
needed} may well have further popularize NumPy as a tool of choice for
modern data science.

As an open source project, NumPy also did not cost money, and was
unencumbered by license servers, dongles, and the likes.

In terms of \emph{philosophical} appeal, NumPy is open source
software.  So, of course, are many other packages that are no longer
relevant, and by itself it does not motivate NumPy's sucess. However,
many developers would never have participated in building a closed
project.  Open source software also appealed the many proponents of
reproducible science, who wanted a transparent, introspectable stack
of software underlying their results.

\emph{Socially}, there was a sense of building something
impactful together, for the benefit of many others.  Participating in
such an endeavor, within a welcoming community of like-minded
indivuals, held a powerful attraction for many early contributors.

Throughout its development, NumPy has upheld strong \emph{technical}
principles.  It is a community of practice that enshrines a culture of
employing software engineering practice to improve collaboration and
reduce error \cite{millman2014developing}.  This culture is not only
adopted by leaders in the project, but also enthusiastically taught to
newcomers. The NumPy team was early in adopting distributed revision
control and code review to improve collaboration on code, and
continuous testing that runs a large battery of automated tests for
every proposed change to NumPy.  The project has comprehensive,
high-quality documentation, integrated with the source
code \cite{vanderwalt2008scipy,harrington2008scipy,harrington2009scipy}.

@@

This culture of using best practices for producing reliable scientific software
has been eagerly adopted by the ecosystem of libraries that build on NumPy.
For example, in a recent award given by the Royal Astronomical Society to
Astropy, they state:
\begin{quotation}
\noindent\emph{The Astropy Project has provided hundreds of junior scientists
with experience in professional-standard software development practices
including use of version control, unit testing, code review and issue tracking
procedures. This is a vital skillset for modern researchers that is often
missing from formal university education in physics or astronomy.}
\end{quotation}
Community members explicitly work to address this lack of formal education
through formal courses and workshops
\cite{wilson-software-carpentry,hannay-scientific-software-survey,millman2018teaching}.
%\fixme{cite https://scipy-school.org/}.

Lastly, NumPy benefited greatly from the simplicity of its underlying data
structure, which made it a natural standard for exchanging array data between
libraries.  It also made it easy for other libraries to develop fast and
memory-efficient compiled code, usually in C or Fortran, that could manipulate
these arrays and pass them back to Python.

Over time the role of the NumPy has changed. It is no longer merely an
array library, but also a standard API for tensor computation and a
central coordinating mechanism between array types and technologies.
NumPy has become a fully-fledged array programming paradigm, that
continues to drive numerical exploration and computation in scientific
Python.

% S:
%
% Thinking about enabling factors, I'd characterize them into three categories:
%
% - Practical
% - Philosophical
% - Social
% - Technical
%
% E.g., practical: reason why we use Python (learn one language for
% everything); students don't have money, so want to avoid impracticalities of
% license dongles/servers, etc.
%
% Philosophical: science should be open, transparent; our software should be
% controlled by scientists, not designers that we don't have access to
%
% Social: joy of building these things together, friendly welcome into the
% community for many of us---appreciation of our work and hours

>>>>

Exposing array programming primitives, as well as the surrounding ecosystem of
tools, in Python---an interpreted language---creates an ideal environment for
interactive, exploratory data analysis where users may iteratively inspect,
manipulate, and visualize their data.

IPython \cite{perez2007ipython}

Jupyter \cite{Kluyver:2016aa}

@@

@@

NumPy has subsequently seen prolific adoption, with millions of users,
first from research and later also industry.
It spurred the development of the larger scientific Python ecosystem and
heralded the current era of wide-spread use of Python for scientific computing.
The reasons for its success are complex, but include the fact NumPy
(a) is built on top of Python, (b) has a strong culture of collaboration
and good software development practices, and (c) provides one of the
best array programming environments for scientific data analysis.

\emph{Python} is an open source, general-purpose, interpreted programming language
well-suited to standard programming tasks such as cleaning data,
interacting with web resources, and parsing text.
Adding fast array operations and linear algebra allows scientist to do all
their work within a single language---and one that has the advantage of
being famously easy to learn and teach.
In fact, Python's adoption as
a primary learning language in many universities (e.g., many lower division
UC Berkeley courses use Python including Foundations of Data Science,
The Beauty and Joy of Computing,
Introduction to Computational Thinking with Data,
and The Structure and Interpretation of Computer Programs) has
further popularized NumPy as a tool of choice for modern data science.

Even though NumPy is not part of Python's standard library,
we have a health relationship to the core development team.
\fixme{we've gotten a number of features: slicing syntax, implicit tuples,
complex numbers, buffer protocol, matrix multiplication operator}
\fixme{but we benefit from separate governance}

NumPy is a community of practice with a strong \emph{culture} of
employing software engineering practice to improve collaboration and
reduce error \cite{millman2014developing}.  This culture is not only
adopted by leaders in the project, but also enthusiastically taught to
newcomers. The NumPy team was early in adopting distributed revision
control and code review to improve collaboration on code, and
continuous testing that runs a large battery of automated tests for
every proposed change to NumPy.  The project has comprehensive,
high-quality documentation, integrated with the source
code \cite{vanderwalt2008scipy,harrington2008scipy,harrington2009scipy}.

@@

This culture of using best practices for producing reliable scientific software
has been eagerly adopted by the ecosystem of libraries that build on NumPy.
For example, in a recent award given by the Royal Astronomical Society to
Astropy, they state:
\begin{quotation}
\noindent\emph{The Astropy Project has provided hundreds of junior scientists
with experience in professional-standard software development practices
including use of version control, unit testing, code review and issue tracking
procedures. This is a vital skill set for modern researchers that is often
missing from formal university education in physics or astronomy.}
\end{quotation}
Community members explicitly work to address this lack of formal education
through formal courses and workshops
\cite{wilson-software-carpentry,hannay-scientific-software-survey,millman2018teaching}.

The \emph{array programming} paradigm originated with APL in the 50s and 60s.
It is characterized by working interactively with array data using
vectorized operations.
Its draw for scientist is captured by the following description of
how APL users work \cite{kromberg2007apl}:
\begin{quote}
Users of APL tend to ``live in their data''. Interactive interpreters allow them
to inspect, almost feel their  way forward, discovering successful snippets
of code through experimentation and collecting them
into imperative or functional programs according to taste.
\end{quote}
Scientist exploring data at the IPython command prompt with the
whole scientific Python ecosystem available to them often work in
this way.  This type of interactive scientific exploration is
increasingly taking place in Jupyter notebooks.

NumPy combines the performance of C, the expressive power of array
programming, and the Python programming language in a mature
well-tested and well-documented community developed library.
It builds on decades of work refining the array programming
syntax from APL through Yorick and others.
It also made it easy for other libraries to develop fast and
memory-efficient compiled code, usually in C or Fortran, that could manipulate
these arrays and pass them back to Python.
Popular libraries increased the user community and that
in turn led to new libraries adding additional functionality
to the scientific Python ecosystem in a virtuous cycle.

Over time the role of the NumPy has evolved. It is no longer just the
foundational array library underlying the scientific Python ecosystem,
but it is also a standard API for tensor computation and a
central coordinating mechanism between array types and technologies.
